A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location

### 2025-05-20 manuscript.qmd
new uncertainty estimation

1. In the **E**xpectation step, we compute the responsibilities $\gamma(\epsilon_k^m)$, which represent the probabilities of a patient $k$ originating from one of the models $m$, given the current estimates of $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$, as given in @eq-responsibilities.
2. During the **M**maximization step, we find a new set of parameters that maximize the new expected complete data likelihood (@eq-expected-complete-likelihood). For the mixture coefficients, we can even find an analytic solution to the new maximum:
   $$
   \pi_m^s = \frac{1}{|K_s|} \sum_{k \in K_s} \gamma(\epsilon_k^m)
   $$
   where we sum over the set of all patients $K_s$ with their tumor in subsite $s$.\
   The models' new parameters are found by numerically maximizing the respective likelihood, weighted by the responsibilities:
   $$
   \ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_k^K \gamma(\epsilon^m_{k}) \big[ \ln \pi_m^{s_k} + \ln P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right) \big]
   $$ {#eq-maximization-step}

@@

>>>>

1. In the **E**xpectation step, we compute the responsibilities $\gamma(\epsilon_k^m)$, which represent the probabilities of a patient $k$ originating from one of the models $m$, given the current estimates of $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$, as given in @eq-responsibilities.
2. During the **M**aximization step, we find a new set of parameters that maximize the new expected complete data likelihood (@eq-expected-complete-likelihood). For the mixture coefficients, we can even find an analytic solution to the new maximum:
   $$
   \pi_m^s = \frac{1}{|K_s|} \sum_{k \in K_s} \gamma(\epsilon_k^m)
   $$
   where we sum over the set of all patients $K_s$ with their tumor in subsite $s$.\
   The models' new parameters are found by numerically maximizing the respective likelihood, weighted by the responsibilities:
   $$
   \ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_k^K \gamma(\epsilon^m_{k}) \big[ \ln \pi_m^{s_k} + \ln P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right) \big]
   $$ {#eq-maximization-step}

@@

## Uncertainty estimation

To estimate the uncertainty associated with both the model parameters and the resulting risk estimates, we employ a bootstrapping approach. Although more comprehensive sampling methods exist for the EM algorithm—such as the imputation-posterior (IP) algorithm, which samples from both model parameters and latent variables—we opt for a computationally more efficient alternative that still yields multiple estimates of all parameters.

Our resampling procedure involves the following steps:

1. Resample the observed data with replacement, maintaining the original number of samples per ICD code.

2. Fit the model to each resampled dataset.

3. Extract the model parameters and risk estimates from each fitted model.

This method enables a straightforward uncertainty analysis while avoiding the computational cost of more intensive sampling techniques.
